\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing economic inequality with tax data - Switzerland from 1945 to 2010}

### 2014-07-02 paper/subdocuments/introduction.Rnw
introduced figure environments

<<Swiss inequality Plot>>=
library(ggplot2)
library(foreign)
library(reshape, quietly=TRUE, warn.conflicts = FALSE)
number_ticks <- function(n) {function(limits) pretty(limits, n)}
swissGini<-read.csv("../../data/swissGini.csv",header=TRUE,sep=";")
swissGini_long<-melt(swissGini,id="Year")
ggplot(data=swissGini_long,
  aes(y=value,x=Year),shape=variable)+
  ylab("Gini of equivalised disposable income") +
  scale_y_continuous(limits=c(0.18,0.4)) +
  scale_x_continuous(limits=c(1982,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=variable),size=3)+
  geom_line(data=swissGini_long[!is.na(swissGini_long$value),],aes(linetype=variable))+
  theme_bw()
@

>>>>

<<Swiss inequality Plot, fig.cap='',fig.pos='H'>>=
library(ggplot2)
library(foreign)
library(reshape, quietly=TRUE, warn.conflicts = FALSE)
number_ticks <- function(n) {function(limits) pretty(limits, n)}
swissGini<-read.csv("../../data/swissGini.csv",header=TRUE,sep=";")
swissGini_long<-melt(swissGini,id="Year")
ggplot(data=swissGini_long,
  aes(y=value,x=Year),shape=variable)+
  ylab("Gini of equivalised disposable income") +
  scale_y_continuous(limits=c(0.18,0.4)) +
  scale_x_continuous(limits=c(1982,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=variable),size=3)+
  geom_line(data=swissGini_long[!is.na(swissGini_long$value),],aes(linetype=variable))+
  theme_bw()
@

### 2014-07-02 paper/subdocuments/results.Rnw
introduced figure environments

\emph{Gini coefficients for taxable income, income before deductions and taxable income after fedaral tax}

@@

<<different_ginis>>=
df$G_reink[df$G_reink==1]<-NA
df$G_taxed[df$G_taxed==1]<-NA
library(reshape)
number_ticks <- function(n) {function(limits) pretty(limits, n)}
dflong <- melt(df[,c("G_steink","G_reink","G_taxed","kanton","steuerperiode")], id=c("kanton","steuerperiode"))
names(dflong)[3]<-"Income"
names(dflong)[4]<-"Gini"
levels(dflong$Income)<-c("taxable income","income before deductions","taxable income after tax")
ggplot(dflong[dflong$kanton=="CH",], aes(x=steuerperiode,y=Gini,shape=Income))+
  scale_y_continuous(limits=c(0.3,0.55))+
  scale_x_continuous(limits=c(1982,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=Income),size=3)+
  geom_line(data=dflong[dflong$kanton=="CH",],aes(linetype=Income))+
  theme_bw()+
  xlab("tax period")+
  theme(legend.justification=c(0,1),legend.position=c(0, 1))
#ggplot(dflong, aes(x=steuerperiode,y=Gini,colour=Type))+geom_line()+facet_wrap(~kanton)+theme_bw()+xlab("tax period")
@

@@

<<equivalencescale>>=
library(ggplot2)
number_ticks <- function(n) {function(limits) pretty(limits, n)}
ggplot(bdlong, aes(x=Jahr,y=Gini,shape=Income))+
  scale_y_continuous(limits=c(0.3,0.55))+
  scale_x_continuous(limits=c(1970,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=Income),size=3)+
  geom_line(data=bdlong,aes(linetype=Income))+
  theme_bw()+
  xlab("tax period")+
  theme(legend.justification=c(0,1),legend.position=c(0, 1))
@

@@

<<reldist20032010, echo=FALSE>>=
library(reldist, quietly=TRUE, warn.conflicts = FALSE)

@@

The FTA stopped to publicly report data for special cases after tax period 1993/94. 
For more recent periods however we can compare already calculated percentiles provided by the FTA\footnote{These calculations were done on commission of the FTA within the SNF project Sinergia Nr. 130648 "The Swiss Confederation: A Natural Laboratory for Research on Fiscal and Political Decentralization" by Raphael Parchet and Stefanie Brilon in coordination with Prof. Dr. Marius Brülhart.} which are based on both special and normal cases to normal cases alone.
We will have a look at both, the 1993/94 period as the last period where both 
numbers were reported as well as 2010 where we compare data from the same source (FTA)
but based on different reporting (aggregated data vs. percentiles).

@@

<<specialcases9394>>=
#plot relative distribution

@@

1993/94 including special cases changes the distribution slightly at both ends. Special cases have a slightly lower median income and therefore are more frequent in the lower percentiles and less frequent in the upper percentiles. In addition there is an effect of shape: the distribution of special cases is more skewed so that lower percentiles show lower incomes and higher percentiles show higher incomes compared to just normal cases.

@@

<<specialcases10>>=

@@

2010 the picture is similar but more apparent: There is a small effect of median shift lowering incomes in the lowest percentiles. The effect of shape however contributes most.

@@

<<with_without_zeros_Gini,warning=FALSE>>=
df <- read.csv("../../data/ginis_und_perzentile_normal.csv",sep="\t")
library(reshape)
dflong <- melt(df[,c("G_steink","G_steink0","G_steink_halb","kanton","steuerperiode")], id=c("kanton","steuerperiode"))
names(dflong)[3]<-"Type"
names(dflong)[4]<-"Gini"
levels(dflong$Type)<-c("without zeros","including zeros","imputed values for zeros")
ggplot(dflong[dflong$kanton=="CH" & dflong$steuerperiode>=1995.5,], 
       aes(x=steuerperiode,y=Gini,shape=Type,linetype=Type))+
  scale_y_continuous(limits=c(0.18,0.6))+
  scale_x_continuous(limits=c(1994,2012),breaks=number_ticks(10)) +
  geom_line(size=0.5)+
  geom_point(aes(shape=Type),size=3)+
  theme_bw()+
  xlab("tax period")+
  theme(legend.justification=c(0,1),legend.position=c(0, 1))

@@

<<with_without_zeros_RelDist_mitBruelhart,warning=FALSE>>=
bd.2010 <- filter(read.csv("../../data/data_Schweiz.csv", header=TRUE,sep=";"),Veranlagungsperiode==2010,Einheit=="Total")
bd.0.2010 <- filter(read.csv("../../data/data_Schweiz_mitNull.csv", header=TRUE,sep=";"),Veranlagungsperiode==2010,Einheit=="Total")
start <- which(names(bd.2010)=="p1")
end <- which(names(bd.2010)=="p99_99")
percentile <- c(1,5,10,20,25,30,40,50,60,70,75,80,90,95,96,97,98,99,99.5,99.9,99.99)
bd.2010<- data.frame(t(bd.2010[1,start:end]),percentile)
names(bd.2010)[1]=c("inc")
bd.0.2010<- data.frame(t(bd.0.2010[1,start:end]),percentile)
names(bd.0.2010)[1]=c("inc")

@@

<<with_without_zeros_RelDist_mitESTV,warning=FALSE>>=

@@

<<BruelhartvsHABE,warning=FALSE>>=
library(ggplot2)
library(reldist, quietly=TRUE, warn.conflicts = FALSE)

@@

<<benplot>>=
 ### libs
library(reshape, quietly=TRUE, warn.conflicts = FALSE)
library(foreign)
library(dplyr)
library(reldist)
library(ggplot2)

>>>>

\emph{Gini coefficients for taxable income, income before deductions and taxable income after federal tax}

@@

<<different_ginis, fig.cap='',fig.pos='H'>>=
df$G_reink[df$G_reink==1]<-NA
df$G_taxed[df$G_taxed==1]<-NA
library(reshape)
number_ticks <- function(n) {function(limits) pretty(limits, n)}
dflong <- melt(df[,c("G_steink","G_reink","G_taxed","kanton","steuerperiode")], id=c("kanton","steuerperiode"))
names(dflong)[3]<-"Income"
names(dflong)[4]<-"Gini"
levels(dflong$Income)<-c("taxable income","income before deductions","taxable income after tax")
ggplot(dflong[dflong$kanton=="CH",], aes(x=steuerperiode,y=Gini,shape=Income))+
  scale_y_continuous(limits=c(0.3,0.55))+
  scale_x_continuous(limits=c(1982,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=Income),size=3)+
  geom_line(data=dflong[dflong$kanton=="CH",],aes(linetype=Income))+
  theme_bw()+
  xlab("tax period")+
  theme(legend.justification=c(0,1),legend.position=c(0, 1))
#ggplot(dflong, aes(x=steuerperiode,y=Gini,colour=Type))+geom_line()+facet_wrap(~kanton)+theme_bw()+xlab("tax period")
@

@@

<<equivalencescale, fig.cap='',fig.pos='H'>>=
library(ggplot2)
number_ticks <- function(n) {function(limits) pretty(limits, n)}
ggplot(bdlong, aes(x=Jahr,y=Gini,shape=Income))+
  scale_y_continuous(limits=c(0.3,0.55))+
  scale_x_continuous(limits=c(1970,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=Income),size=3)+
  geom_line(data=bdlong,aes(linetype=Income))+
  theme_bw()+
  xlab("tax period")+
  theme(legend.justification=c(0,1),legend.position=c(0, 1))
@

@@

<<reldist20032010, echo=FALSE, fig.cap='',fig.pos='H'>>=
library(reldist, quietly=TRUE, warn.conflicts = FALSE)

@@

The FTA distinguishes normal from special cases as described in the data section. To test whether it matters which cases the researcher looks at we want to compare the distributions of normal and special cases. Unfortunately, the FTA stopped to publicly report data for special cases after tax period 1993/94. Therefore we will compare the two distributions for a rather old dataset. However the FTA does report aggregate statistics (e.g. percentiles) based on a pool of all cases (normal and special) for more recent periods\footnote{These calculations were done on commission of the FTA within the SNF project Sinergia Nr. 130648 "The Swiss Confederation: A Natural Laboratory for Research on Fiscal and Political Decentralization" by Raphael Parchet and Stefanie Brilon in coordination with Prof. Dr. Marius Brülhart.} which allows us to do a corresponding analysis for 2010 as well.

@@

<<specialcases9394, fig.cap='',fig.pos='H'>>=
#plot relative distribution

@@

1993/94 a pooled data set of normal and special cases has a slightly higher density at both ends compared to data based in normal cases only. Special cases appear to have a slightly lower median income and their distribution is more skewed. Therefore special cases are more polarized than normal cases, i.e. striving away from the median (positive Median Index of 0.02). This tendency is more pronounces in the lower than upper part  of the distribution (Lower Index of 0.029 compared to Upper Index if 0.01). 
As the special cases consist of a broad mix of individuals it remains unclear which factor explains the differences of both distributions. Possible explanations can be immigrants partly concentrating in lower income percentiles, low income artists who belong to the special cases or a more technical selection effect: tax units not liable for taxation throughout the whole year are special cases; those cases might have lower incomes, e.g. if they moved and stopped working. To get a more complete picture we can look how the two distributions relate to each other in 2010 (see FIGURE SPECIAL2010).

@@

<<specialcases10, fig.cap='',fig.pos='H'>>=

@@

2010 the picture is similar but more apparent: Special cases appear more frequent around the lower percentiles of the pooled distribution (Lower Index of 0.039), however 2010 there is a more noteworthy effect in the upper part of the distribution (Upper Index of 0.022). According to figure \ref{fig:specialcases10} we can attribute this effect to the top percentiles. This gives credibility to the thesis that rich immigrants whose number increased between 1994 and 2010 drive the effect.

@@

<<with_without_zeros_Gini,warning=FALSE, fig.cap='',fig.pos='H'>>=
df <- read.csv("../../data/ginis_und_perzentile_normal.csv",sep="\t")
library(reshape)
dflong <- melt(df[,c("G_steink","G_steink0","G_steink_halb","kanton","steuerperiode")], id=c("kanton","steuerperiode"))
names(dflong)[3]<-"Type"
names(dflong)[4]<-"Gini"
levels(dflong$Type)<-c("without zeros","including zeros","imputed values for zeros")
ggplot(dflong[dflong$kanton=="CH" & dflong$steuerperiode>=1995.5,], 
       aes(x=steuerperiode,y=Gini,shape=Type,linetype=Type))+
  scale_y_continuous(limits=c(0.18,0.6))+
  scale_x_continuous(limits=c(1994,2012),breaks=number_ticks(10)) +
  geom_line(size=0.5)+
  geom_point(aes(shape=Type),size=3)+
  theme_bw()+
  xlab("tax period")+
  theme(legend.justification=c(0,1),legend.position=c(0, 1))

@@

<<with_without_zeros_RelDist_mitBruelhart,warning=FALSE, fig.cap='',fig.pos='H'>>=
bd.2010 <- filter(read.csv("../../data/data_Schweiz.csv", header=TRUE,sep=";"),Veranlagungsperiode==2010,Einheit=="Total")
bd.0.2010 <- filter(read.csv("../../data/data_Schweiz_mitNull.csv", header=TRUE,sep=";"),Veranlagungsperiode==2010,Einheit=="Total")
start <- which(names(bd.2010)=="p1")
end <- which(names(bd.2010)=="p99_99")
percentile <- c(1,5,10,20,25,30,40,50,60,70,75,80,90,95,96,97,98,99,99.5,99.9,99.99)
bd.2010<- data.frame(t(bd.2010[1,start:end]),percentile)
names(bd.2010)[1]=c("inc")
bd.0.2010<- data.frame(t(bd.0.2010[1,start:end]),percentile)
names(bd.0.2010)[1]=c("inc")

@@

<<with_without_zeros_RelDist_mitESTV,warning=FALSE, fig.cap='',fig.pos='H'>>=

@@

<<BruelhartvsHABE,warning=FALSE, fig.cap='',fig.pos='H'>>=
library(ggplot2)
library(reldist, quietly=TRUE, warn.conflicts = FALSE)

@@

<<benplot, fig.cap='',fig.pos='H'>>=
 ### libs
library(reshape, quietly=TRUE, warn.conflicts = FALSE)
library(foreign)
library(dplyr)
library(reldist)
library(ggplot2)
